---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
version <- as.character(packageVersion("microbialkitchen"))
```

# Microbial Kitchen (formerly MediaChemTools)

[![R-CMD-check](https://github.com/KopfLab/microbialkitchen/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/KopfLab/microbialkitchen/actions/workflows/R-CMD-check.yaml)
[![Documentation](https://img.shields.io/badge/docs-online-green.svg)](https://microbialkitchen.kopflab.org/reference/)
[![codecov](https://codecov.io/github/KopfLab/microbialkitchen/branch/master/graphs/badge.svg)](https://codecov.io/github/Kopflab/microbialkitchen) 

## About

The **[microbialkitchen](https://microbialkitchen.kopflab.org/)** package is a collection of tools to simplify working with the chemical composition and speciation of defined culture media for microbial physiology and environmental microbiology research. It includes a wide range of general purpose functionality for chemical applications including built-in, data-frame-compatible [chemical quantities](https://microbialkitchen.kopflab.org/articles/quantities.html) (volume, mass, molarity, temperature, pressure, etc.) that automatically keep track of their units and metric scaling, as well as more specialized tools for the assembly and comparison of culturing media recipes, pH buffering strategies and aqueous speciation. All basic data types and operations are fully implemented, documented and ready to use. However, since the package is still in active development and some syntax and function names may change.

## Installation

You can install **[microbialkitchen](https://microbialkitchen.kopflab.org/)** from github with the devtools package. 

```{r gh-installation, eval = FALSE}
# install.packages("devtools") 
devtools::install_github("KopfLab/microbialkitchen")
```

## Functionality

 - [function reference](https://microbialkitchen.kopflab.org/reference/) for details on all functions
 - [quantities vignette](https://microbialkitchen.kopflab.org/articles/quantities.html) on built-in chemical quantities 
 - [operations vignette](https://microbialkitchen.kopflab.org/articles/operations.html) on working with quantities
 - carbonate chemistry vignettes for working with carbonate buffered media systems:
    - [equations vignette](https://microbialkitchen.kopflab.org/articles/carbonate_chemistry_equations.html)
    - [examples vignette](https://microbialkitchen.kopflab.org/articles/carbonate_chemistry_examples.html)

