---
title: "Carbonate Chemistry Examples"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
  html_document:
    code_folding: show
    df_print: paged
    number_sections: yes
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options:
  chunk_output_type: console
vignette: >
  %\VignetteIndexEntry{Carbonate Chemistry Examples}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

This vignette provides examples of the carbonate chemistry tools implemented by the **microbialkitchen** package to explore carbonate solubility and pH control in open and closed media systems. See the [carbonate chemistry equations vignette](carbonate_chemistry_equations.html) for conceptual background.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse) # load tidyverse tools
library(microbialkitchen) # load microbial kitchen library
```

# CO2 solubility

[See equations](carbonate_chemistry_equations.html#co_2-solubility).

```{r "henry", fig.width=8, fig.height=6}
solubilities <- tibble(
  temperature = seq(from = 0, to = 40, by = 5) %>% qty("C"),
  solubility = calculate_gas_solubility("CO2", temperature)
)

solubilities %>% 
  ggplot() +
  aes(temperature, solubility) +
  geom_line(size = 2) +
  scale_x_qty(unit = "C") +
  theme_bw()
```

# Speciation

[See equations](carbonate_chemistry_equations.html#speciation).

```{r "speciation", fig.width=8, fig.height=6}
# calculate
df_speciation <-
  tibble(
    pH = seq(3, 14, by = 0.1),
    DIC = qty(10, "mM"),
    `H2CO3*` = calculate_carbonic_acid(pH, DIC = DIC),
    HCO3 = calculate_bicarbonate(pH, DIC = DIC),
    CO3 = calculate_carbonate(pH, DIC = DIC)
  )

# visualize
df_speciation %>%
  # pivoting longer for the concentration columns
  pivot_longer(
    names_to = "Species", values_to = "Concentration", 
    cols = c(`H2CO3*`, HCO3, CO3)
  ) %>%
  ggplot() +
  aes(pH, Concentration, color = Species) +
  geom_line() +
  scale_x_continuous(expand = c(0, 0), breaks = c(1:14)) +
  scale_y_qty(each = TRUE) +
  theme_bw()
```


# Open system

[See equations](carbonate_chemistry_equations.html#open-system).

## CO2 only

```{r}
# range of pCO2s
pCO2s <- tibble(
  pCO2 = qty(c(seq(0, 5, by=0.1), seq(5, 20, by=1), seq(20, 200, by=5)), "mbar")
)
# range of temperature
temperatures <- tibble(
  temperature = qty(c(10, 25, 37), "C")
)

# calculate for all combinations
df_pH_vs_pCO2 <-
  crossing(pCO2s, temperatures) %>%
  mutate(
    pH = calculate_open_system_pH(pCO2, temp = temperature),
    DIC = calculate_DIC(pH, pCO2, temp = temperature),
    CO2 = calculate_ideal_gas_molarity(pCO2, temp = temperature)
  )
```

```{r "pH vs pCO2", fig.width = 8, fig.height = 6}
df_pH_vs_pCO2 %>%
  # pivoting longer mixed data types requires explicit units first
  make_qty_units_explicit(DIC = "mM", CO2 = "mM") %>%
  pivot_longer(
    cols = c(pH, `DIC [mM]`, `CO2 [mM]`), 
    names_to = "var", values_to = "value"
  ) %>%
  filter(!(var == "pH" & value > 5.5)) %>%
  mutate(temperature = as_factor(temperature, unit = "C")) %>%
  ggplot() +
  aes(pCO2, value, color = temperature) +
  geom_line() +
  scale_x_qty(expand = c(0, 0)) +
  scale_color_brewer(palette = "Set1") +  
  facet_grid(var~., scales = "free_y") +
  expand_limits(x = 0) +
  theme_bw() + 
  labs(y = NULL)
```

## Adjusting pH with alkalinity

```{r}
# to calculate how much base (NaOH or NaHCO3) to add
calculate_open_system_alkalinity(pH = 6.8, pCO2 = qty(0.4, "mbar"))
calculate_open_system_alkalinity(pH = 6.8, pCO2 = qty(50, "mbar"))
calculate_open_system_alkalinity(pH = 6.8, pCO2 = qty(200, "mbar"))

# to calculate pH with addition of a specific amount of base
calculate_open_system_pH(pCO2 = qty(0.4, "mbar"), alkalinity = qty(5, "mM"))
calculate_open_system_pH(pCO2 = qty(50, "mbar"), alkalinity = qty(5, "mM"))
calculate_open_system_pH(pCO2 = qty(200, "mbar"), alkalinity = qty(5, "mM"))

# to calculate pH with addition of a specific amount of acid
calculate_open_system_pH(pCO2 = qty(0.4, "mbar"), alkalinity = qty(-1, "nM"))
calculate_open_system_pH(pCO2 = qty(50, "mbar"), alkalinity = qty(-1, "nM"))
calculate_open_system_pH(pCO2 = qty(200, "mbar"), alkalinity = qty(-1, "nM"))
```

### Visualization

```{r "unbalanced ions for pH", fig.width = 9, fig.height = 6}
# range of pCO2s
pCO2s <- tibble(pCO2 = qty(c(0.4, 50, 200), "mbar"))
# range of temperature
temperatures <- tibble(temperature = qty(c(25, 37), "C"))

# calculate for all combinations
zero <- qty(0, "mM")
df_base_vs_pH <-
  crossing(pCO2s, temperatures, pH = seq(3, 7.5, by = 0.05)) %>%
  mutate(
    ions = calculate_open_system_alkalinity(pH, pCO2, temp = temperature),
    `Add base (e.g. NaOH)` = case_when(ions > zero ~ ions, TRUE ~ zero),
    `Add acid (e.g. HCl)` = case_when(ions < zero ~ -1*ions, TRUE ~ zero),
    DIC = calculate_DIC(pH, pCO2, temp = temperature)
  )

p <- df_base_vs_pH %>%
  pivot_longer(
    cols = c(`Add base (e.g. NaOH)`, `Add acid (e.g. HCl)`, `DIC`),
    names_to = "var", values_to = "concentration"
  ) %>% 
  filter(concentration > zero) %>%
  mutate(
    temperature = as_factor(temperature, unit = "C"),
    panel = as_factor(pCO2, unit = "mbar")
  ) %>%
  ggplot() +
  aes(pH, concentration, color = var, linetype = temperature) +
  geom_vline(xintercept = 7, color = "black", size = 1) +
  geom_line() +
  scale_x_continuous(breaks = 1:14, expand = c(0, 0)) +
  scale_y_qty(each = TRUE) +
  scale_color_brewer("Component", palette = "Set1") +  
  facet_wrap(~panel, nrow = 1, scales = "free_y") +
  theme_bw()
p
```

### Log scale

```{r "unbalanced ions for pH - log scale", fig.width = 9, fig.height = 6}
# easier to see in log space and equal y axis + a vertical line
p_log <- p +
  scale_y_qty(each = TRUE, trans = "log10") +
  facet_wrap(~panel, nrow = 1)
p_log
```

## Adjusting pH with a buffer

For simplicty keeping temperature constant at the default (25C).

```{r}
buffer <- qty(c(1, 10, 50), "mM")
# to calculate how much base (NaOH or NaHCO3) to add
df <- tibble(
  buffer = buffer,
  add = calculate_open_system_alkalinity(
    pH = 6.8,
    pCO2 = qty(50, "mbar"),
    buffer = buffer,
    buffer_pKa = 7.5
  )
)
df

# what if buffer is mono-sodium? (negative means HCl or other strong acid required instead of base)
df %>% mutate(add_if_salt_buffer = add - buffer)

# to calculate pH with addition of a specific buffer
df %>%
  mutate(
    pH = calculate_open_system_pH(
      pCO2 = qty(50, "mbar"),
      buffer = buffer,
      alkalinity = add,
      buffer_pKa = 7.5
    )
  )
```

### Different concentrations of weak acid (same pKa)

```{r "unbalanced ions for pH with buffer", fig.width = 9, fig.height = 6}
# range of pCO2s
pCO2s <- tibble(pCO2 = qty(c(0.4, 50, 200), "mbar"))
# range of buffers
buffers <- tibble(
  buffer = qty(c(0, 20, 20, 20, 50), "mM"),
  buffer_pKa = c(7.5, 7.5, 9, 6, 7.5)
)

# calculate for all combinations
df_base_w_buffer_vs_pH <-
  crossing(pCO2s, buffers, pH = seq(3, 7.5, by = 0.05)) %>%
  mutate(
    ions = calculate_open_system_alkalinity(pH, pCO2, buffer = buffer, buffer_pKa = buffer_pKa),
    `Add base (e.g. NaOH)` = case_when(ions > zero ~ ions, TRUE ~ zero),
    `Add acid (e.g. HCl)` = case_when(ions < zero ~ -1*ions, TRUE ~ zero),
    DIC = calculate_DIC(pH, pCO2)
  )

# visualize
plot_df <- df_base_w_buffer_vs_pH %>%
  pivot_longer(
    cols = c(`Add base (e.g. NaOH)`, `Add acid (e.g. HCl)`, `DIC`),
    names_to = "var", values_to = "concentration"
  ) %>% 
  mutate(
    panel = as_factor(pCO2, unit = "mbar"),
    Buffer = sprintf("%s (pkA=%s)", buffer, buffer_pKa)
  ) %>%
  filter(concentration > zero)

p_log %+% 
  filter(plot_df, buffer_pKa == 7.5) %+%
  aes(linetype = Buffer) +
  labs(title = "Different concentrations of a weak acid buffer")
```

### Same concentration of weak acid (different pKa)

```{r "unbalanced ions for pH with buffer difference pKa", fig.width = 9, fig.height = 6}
p_log %+% 
  filter(plot_df, buffer == qty(20, "mM")) %+%
    aes(linetype = Buffer) +
    labs(title = "Same concentration of a weak acid buffer with different pKa")
```

### Different concentration of salt buffer (same pKa)

If the buffer is a salt instead of an acid, it contributes to the alkalinity.

```{r "unbalanced ions for pH with different buffer concs", fig.width = 9, fig.height = 6}
plot_df2 <-
  crossing(pCO2s, buffers, pH = seq(3, 7.5, by = 0.05)) %>%
  mutate(
    ions = calculate_open_system_alkalinity(pH, pCO2, buffer = buffer, buffer_pKa = buffer_pKa) - buffer,
    `Add base (e.g. NaOH)` = case_when(ions > zero ~ ions, TRUE ~ zero),
    `Add acid (e.g. HCl)` = case_when(ions < zero ~ -1*ions, TRUE ~ zero),
    DIC = calculate_DIC(pH, pCO2)
  ) %>%
  pivot_longer(
    cols = c(`Add base (e.g. NaOH)`, `Add acid (e.g. HCl)`, `DIC`),
    names_to = "var", values_to = "concentration"
  ) %>% 
  mutate(
    panel = as_factor(pCO2, unit = "mbar"),
    buffer = sprintf("%s (pkA=%s)", buffer, buffer_pKa)
  ) %>%
  filter(concentration > zero)

p_log %+% 
  filter(plot_df2, buffer_pKa == 7.5) %+%
  aes(linetype = buffer) +
  labs(title = "Different concentrations of a mono-valent salt buffer")
```

## pH difference upon atmosphere switching

A rare environmental but common laboratory scenario is the sometimes unexpected change in pH upon shifting from standard atmospheric CO2 (400ppm, let's face reality...) to some artifical higher CO2 atmosphere without adjusting anything else.

```{r, "pH change vs pCO2 and acid base with buffer", fig.width = 8, fig.height = 7, warning=FALSE}
df_pH_switch <-
  df_base_w_buffer_vs_pH %>%
  filter(get_qty_value(pCO2, "mbar") == 0.4) %>%
  rename(init_pCO2 = pCO2) %>%
  crossing(pCO2 = qty(c(50, 200), "mbar")) %>%
  mutate(
    `new pH` = calculate_open_system_pH(
      pCO2 = pCO2, buffer = buffer, buffer_pKa = buffer_pKa, alkalinity = ions
    ),
    `pH difference` = `new pH` - pH,
    panel = as_factor(pCO2, unit = "mbar")
  ) %>%
  pivot_longer(
    cols = c(`new pH`, `pH difference`),
    names_to = "var", values_to = "value", 
  )

p %+% 
  mutate(df_pH_switch, buffer = as_factor(buffer)) %+%
  aes(y = value, color = buffer, linetype = as_factor(buffer_pKa)) +
  facet_grid(var~panel, scales = "free_y") +
  scale_color_brewer(palette = "Set1") +  
  scale_linetype_manual(values = c(2, 1, 3)) +
  labs(linetype = "pKa", y = "")
```

# Closed system

[See equations](carbonate_chemistry_equations.html#closed-system).

## CO2 consumption

Scenario: media is equilibrated with an initial CO2 pressure and adjusted to a specific starting pH (which requires the appropriate addition of alkalinity) in presence or absence of a pH buffer. Then the vessel is closed off and has a starting total inorganic carbon from its headspace and liquid that can then be consumed. Here a few parameters are explored but additional variation will arise from changes in buffer concentration, pKa and the proportion of headspace to liquid.

```{r "CO2 consumption", message = FALSE, fig.width = 9, fig.height=8}
df_CO2_consumption <-
  crossing(
    tibble(
      # initial medium
      pH_start = 7,
      pCO2_start = qty(c(0.4, 50, 200), "mbar")
    ),
    tibble(
      # buffer
      buffer = qty(c(0, 50), "mM"),
      pKa = 7.5
    )
  ) %>%
  mutate(
    # required alkalinity for pH
    pH_start = 7,
    alkalinity = calculate_open_system_alkalinity(
      pH = pH_start, pCO2 = pCO2_start,
      buffer = buffer, buffer_pKa = pKa),
    # close off
    V_liquid = qty(10, "mL"),
    V_gas = qty(10, "mL"),
    TIC_start = calculate_closed_system_TIC(pH = pH_start, pCO2 = pCO2_start, V_liquid = V_liquid, V_gas = V_gas)
  ) %>%
  # range of TIC used up (%)
  crossing(TIC_used = seq(0, 1, by = 0.01)) %>%
  # calculation of system at each point
  mutate(
    TIC = TIC_start * (1-TIC_used),
    pH = calculate_closed_system_pH(
      TIC = TIC, V_liquid = V_liquid, V_gas = V_gas,
      alkalinity = alkalinity, buffer = buffer, buffer_pKa = pKa),
    pCO2 = calculate_closed_system_pCO2(pH = pH, TIC = TIC, V_liquid = V_liquid, V_gas = V_gas),
    DIC = calculate_DIC(pH, pCO2 = pCO2),
    `H2CO3*` = calculate_carbonic_acid(pH, DIC = DIC),
    HCO3 = calculate_bicarbonate(pH, DIC = DIC),
    CO3 = calculate_carbonate(pH, DIC = DIC)
  ) %>%
   mutate(
    pCO2_start = as_factor(sprintf("%s CO2", pCO2_start)),
    buffer = sprintf("%s (pkA=%s)", buffer, pKa),
  )

# visualize
p_base <-
  df_CO2_consumption %>%
  ggplot() +
  aes(TIC_used, linetype = buffer) +
  scale_y_qty(each = TRUE, trans = "log10", expand = c(0, 0)) +
  scale_x_continuous(
    labels = function(x) sprintf("%.0f%%", 100*x), 
    expand = c(0, 0)) +
  facet_wrap(~pCO2_start) +
  theme_bw() + theme(panel.grid.minor = element_blank()) + 
  labs("TIC consumed")

p_IC <- p_base %+%
  aes(y = concentration, color = species) +
  geom_line(
    data = function(df)
      df %>% pivot_longer(
        cols = c(DIC, `H2CO3*`, HCO3, CO3),
        names_to = "species",
        values_to = "concentration",
      ) %>% filter(concentration > qty(1, "nM"))
  )

p_pH <- p_base %+% 
  aes(y = pH, color = NULL) +
  geom_line() +
  scale_y_continuous() +
  theme(legend.position = "none")

p_CO2 <- p_base %+%
  aes(y = pCO2, color = NULL) +
  geom_line(data = function(df) filter(df, pCO2 > qty(1e-7, "bar"))) +
  theme(legend.position = "none")

library(cowplot)
plot_grid(p_pH, p_IC, p_CO2, ncol = 1, align = "v", axis = "lr")
```


## CO2 production

Scenario: media with organic substrate (e.g. a sugar) is adjusted to an initial starting pH (which requires the appropriate addition of alkalinity) in presence or absence of a pH buffer. Then the vessel is closed off and CO2 is produced from the organic substrate (assuming the entire organic pool is respirable).

```{r "CO2 production", fig.width = 9, fig.height=8}
df_CO2_production <-
  crossing(
    tibble(
      # initial medium
      pH_start = 7,
      Corg = qty(c(1, 10, 100), "mM") # respirable carbon molarity
    ),
    tibble(
      # buffer
      buffer = qty(c(0, 50), "mM"),
      pKa = 7.5
    )
  ) %>%
  mutate(
    # required alkalinity for pH adjustment
    pH_start = 7,
    pCO2_start = qty(0.4, "mbar"), # atmospheric
    alkalinity = calculate_open_system_alkalinity(
      pH = pH_start, pCO2 = pCO2_start,
      buffer = buffer, buffer_pKa = pKa),
    # close off
    V_liquid = qty(10, "mL"),
    V_gas = qty(10, "mL"),
    TIC_start = calculate_closed_system_TIC(pH = pH_start, pCO2 = pCO2_start, V_liquid = V_liquid, V_gas = V_gas)
  ) %>%
  # range of organics respired (%)
  crossing(Corg_used = seq(0, 1, by = 0.01)) %>%
  # calculation of system at each point
  mutate(
    TIC = TIC_start + Corg * V_liquid * Corg_used,
    pH = calculate_closed_system_pH(
      TIC = TIC, V_liquid = V_liquid, V_gas = V_gas,
      alkalinity = alkalinity, buffer = buffer, buffer_pKa = pKa),
    pCO2 = calculate_closed_system_pCO2(pH = pH, TIC = TIC, V_liquid = V_liquid, V_gas = V_gas),
    DIC = calculate_DIC(pH, pCO2 = pCO2),
    `H2CO3*` = calculate_carbonic_acid(pH, DIC = DIC),
    HCO3 = calculate_bicarbonate(pH, DIC = DIC),
    CO3 = calculate_carbonate(pH, DIC = DIC)
  ) %>%
  mutate(
    Corg = as_factor(sprintf("%s org. C", Corg)),
    buffer = sprintf("%s (pkA=%s)", buffer, pKa),
  )

# plots
p_IC_b <- p_IC %+% df_CO2_production %+% aes(x = Corg_used) + facet_wrap(~Corg) +
  labs(x = "Organic C consumed")
p_pH_b <- p_pH %+% df_CO2_production %+% aes(x = Corg_used) + facet_wrap(~Corg) +
  labs(x = "Organic C consumed")
p_CO2_b <- p_CO2 %+% df_CO2_production %+% aes(x = Corg_used) + facet_wrap(~Corg) +
  labs(x = "Organic C consumed")

plot_grid(p_pH, p_IC, p_CO2, ncol = 1, align = "v", axis = "lr")
```
