---
title: "Operations"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
  html_document:
    code_folding: show
    number_sections: yes
    df_print: paged
    toc: yes
    toc_depth: 3
    toc_float: yes
editor_options:
  chunk_output_type: console
vignette: >
  %\VignetteIndexEntry{Operations}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r, message=FALSE}
library(microbialkitchen) # load the library
```

```{r, includ=FALSE}
knitr::opts_chunk$set(fig.width = 7, fig.height = 4)
```


# Units & Metric Scaling

All chemical quantities in this package keep track of the units they represent with a base unit as detailed in the [quantities vignette](quantities.html) and a metric prefix. By default, the metric prefix is adjusted automatically to keep numeric values in a range close to 1. However, scaling to a specific prefix is easily achieved.

```{r}
# automatic scaling
qty(5000, "g") # automatically scaled to kg
qty(5000, "g", scale_to_best_metric = FALSE) # stays g
# specific scaling
qty(100, "mg") %>% base_metric() # scale to the base unit (here g)
qty(100, "mM") %>% base_metric() # scale to the base unit (here M)
qty(100, "mg") %>% scale_metric("k")
```

The actual numeric value of a quantity can be retrieved via the `get_qty_value()` function which takes an optional parameter to specify which unit the value should be retrieved in (by default it is the unit that the quantity is currently in). 

```{r}
qty(100, "mg") %>% get_qty_value() # returns 100
qty(0.1, "g") %>% get_qty_value() # returns 100 because the units are "mg"
qty(0.1, "g") %>% get_qty_value("g") # better to be specific upon retrieval
qty(0.1, "g") %>% get_qty_value("kg") # can use any valid unit
qty(0, "C") %>% get_qty_value("F") # or use it for unit conversion
qty(760, "Torr") %>% get_qty_value("atm") # or use it for unit conversion
```

To retrieven numeric values as text that includes their units, simply use `get_qty_text()` instead of `get_qty_value()`. 

```{r}
# get quantity with unit
qty(100, "mg") %>% get_qty_text()
qty(0:10, "C") %>% get_qty_text("F")
qty(760, "Torr") %>% get_qty_text("atm")
```

To retrieve each quantity in it's ideal metric scaling, use `get_qty_text_each()`.

```{r}
# without "each" 
qty(c(0.1, 1, 1000), "g") %>% get_qty_text()

# with "each"
qty(c(0.1, 1, 1000), "g") %>% get_qty_text_each()
```

The units of a quantity object can also be retrieved directly with the `get_qty_units()` function for a single quantity, list of quantities as well as entire data frames (see details on the latter in the [data frames section](#data-frames)).

```{r}
qty(5000, "g") %>% get_qty_units()
x <- list(
  a = qty(5000, "g"),
  b = 42,
  c = qty(100, "mbar")
) 
x %>% get_qty_units()
```

# Arithmetic

Several common arithmetic operations are implemented for easy interconversion between quantities. All arithmetic operations also automatically keep track of the units and metric prefixes for correct calculations.

## Addition and Subtraction

Quantities of the same type (e.g. masses, volumes, etc.) can be added or subtracted with proper interpration of the metric prefixes. The resulting quantity will be scaled to the best metric prefix as described above. Attempts to add or subtract non-matching quantities (e.g. mass + volume) or a quantity and a number without units will fail with an error to avoid unexpect behaviour and ambiguous calculations. 

```{r}
qty(0.003, "g") - qty(2, "mg") + qty(5, "µg") # 1.005 mg
try(qty(1, "g") + qty(1, "L")) # not allowed
try(qty(1, "g") + 1) # not allowed
```

## Multiplication / Division

Quantities can be multipled/divided by a number. The resulting quantity will be scaled to the best metric prefix. This is most commonly used with multiplication/division by 1000.

```{r}
qty(1, "mg") * 1000 # convert mg into g
qty(1, "mg") / 1e6 # convert mg into ng
```

Quantities can also be divided by another quantity of the same type (e.g. a mass by another mass) effectively canceling out the units resulting in a regular number (with the metric prefixes properly taken into consideration).

```{r}
# how many mg in a kg?
qty(1, "mg") / qty(1, "kg")
```

Additional multiplications and divisions are possible for specific combinations of quantities as detailed below. These formulas are each implemented for all three possible arrangements.

### Concentration (Molarity) = Amount / Volume

```{r}
qty(5, "nmol") / qty(50, "mL") # calculation molarity
qty(5, "nmol") / qty(100, "nM") # calculate volume
qty(100, "nM") * qty(50, "mL") # calculate amount
```

### Concentration (Density) = Mass / Volume

```{r}
qty(5, "ng") / qty(50, "mL") # calculate density
qty(5, "ng") / qty(100, "ng/L") # calculate volume
qty(100, "ng/L") * qty(50, "mL") # calculate mass
```

### Amount = Mass / Molecular Mass

```{r}
qty(10, "g") / qty (50, "g/mol") # calculate amount
qty(10, "g") / qty(200, "mmol") # calculate molecular weight
qty(200, "mmol") * qty (50, "g/mol") # calculate mass
```

### Solubility = Molarity / Pressure

```{r}
qty(10, "mM") / qty(200, "mbar") # calculate solubility
qty(10, "mM") / qty(50, "mM/bar") # calculate pressure
qty(50, "mM/bar") * qty (200, "mbar") # calculate molarity
```


## Comparisons

Quantities can be compared with all common logic operators (`>`, `>=`, `<`, `<=`, `==`, `!=`) taking the metric scaling properly into consideration. Attempts to compare non-matching quantities (e.g. mass & volume)  will fail with an error to avoid unexpect behaviour. Comparisons of quantities with numerics are allowed but it is important to be cautious about these since the metric scaling in quantities affects the numeric value.

```{r}
qty(5, "mg") < qty(1, "g")
qty(5, "mg") > qty(10, "ng")
qty(5, "mg") == qty(0.005, "g")
qty(5, "mg") != qty(5, "g")
try(qty(1, "mg") > qty(1, "L")) # not allowed (different quantities)
```

It is important to note that due to machine errors, the `==` is best avoided in favor of more reliable comparisions such as tests that check whether the difference between quantities is smaller than a tiny quantity:

```{r}
(x <- qty(1, "mg"))
(y <- x / 3)
(x2 <- y * 3)
x == x2 # should be identical but is not because of machine precision
abs(x - x2) < qty(1, "fg") # absolute difference smaller than 1 femtogram
```


# Data Frames

Quantities are fully supported in dplyr type data frames (tibbles) and the type and units of a quantity are displayed underneath the column headers, e.g. `V<mL>` to indicate a quantity column that is a volume and has the units of `mL`.

```{r, message=FALSE}
tibble(
  weight = qty(1:5, "mg"),
  volume = qty(20, "mL")
)
```

This also means that all arithmetic operations are allowed within data frames.

```{r}
df <- tibble(
  weight = qty(1:5, "mg"),
  vol = qty(20, "mL"),
  mw = qty(500, "g/mol"),
  amount = weight / mw,
  conc = amount / vol
)
df
```

To get the columns' units use `get_qty_units()`

```{r}
df %>% get_qty_units()
```

To make units of specific columns explicit, use `make_qty_units_explicit()`

```{r}
# make weight and concentration explicit
df %>% make_qty_units_explicit(weight = "g", conc = "mM")

# change the format for the units
df %>% make_qty_units_explicit(weight = "g", conc = "mM", prefix = ".", suffix = "")
```

# Plotting

Quantities are automatically plotted with proper axis labels if used in `ggplot`.

```{r}
# example data set
library(dplyr)
library(tidyr)
df <- 
  crossing(
    conc = qty(c(0.1, 25), "mM"),
    volume = qty(c(0.2, 90, 400), "mL")
  ) %>% 
  mutate(
    amount = conc * volume
  )
df
```

By default, units are included in the x and y axes titles if a quantity is being plotted. Discrete quantities used for other aesthetics can be converted to a factor using `as_factor` which automatically includes the units.

```{r}
library(ggplot2)
df %>% mutate(conc = as_factor(conc)) %>% 
  ggplot() +
  aes(x = volume, y = amount, color = conc) +
  geom_line() +
  facet_wrap(~conc, scales = "free")
```

Alternatively, the unit can be specified using the `scale_x_qty()` and `scale_y_qty()` functions and the `unit` parameter to the `as_factor` function. Units must be valid for the quantities that are plotted.

```{r}
df %>% mutate(conc = as_factor(conc, unit = "M")) %>% 
  ggplot() +
  aes(x = volume, y = amount, color = conc) +
  geom_line() +
  facet_wrap(~conc, scales = "free") +
  scale_x_qty(unit = "L") +
  scale_y_qty(unit = "mmol")
```

It is also possible to use the `each` parameter in `scale_x_qty()` and `scale_y_qty()` as well as the `as_factor` function to use the ideal metric units for each value. For the axes, this switches from including the unit in the axis title to the individual axis values.

```{r}
df %>%
  mutate(conc = as_factor(conc, each = TRUE)) %>% 
  ggplot() +
  aes(x = volume, y = amount, color = conc) +
  geom_line() +
  facet_wrap(~conc, scales = "free") +
  scale_x_qty(each = TRUE) +
  scale_y_qty(each = TRUE)
```

It is also possible to combine `each` and a specific `unit` for the `scale_x_qty()` and `scale_y_qty()` functions. This leads to individual axis value labels in the desired unit.

```{r}
df %>%
  mutate(conc = as_factor(conc, each = TRUE)) %>% 
  ggplot() +
  aes(x = volume, y = amount, color = conc) +
  geom_line() +
  facet_wrap(~conc, scales = "free") +
  scale_x_qty(unit = "L", each = TRUE) +
  scale_y_qty(unit = "mmol", each = TRUE)
```

Lastly, the `scale_x_qty()` and `scale_y_qty()` support all axis transformations supported by `scale_x_continuous()` and `scale_y_continuous()`. For example the `log10` transform can be used like this:

```{r}
df %>%
  mutate(conc = as_factor(conc, each = TRUE)) %>% 
  ggplot() +
  aes(x = volume, y = amount, color = conc) +
  geom_line() +
  facet_wrap(~conc, scales = "free") +
  scale_x_qty(trans = "log10", each = TRUE) +
  scale_y_qty(trans = "log10", each = TRUE)
```

# Concatenation

Quantities can be concatenated using the regular `c()` function (or the more explicit `c_qty()`) as long as they are the same type of quantity (e.g. all masses). Concatenation make sure that the metric prefix is taken into consideration and scales the new vector to the best metric of the median. 

```{r}
c(
  qty(1, "g"), 
  qty(1:3, "mg"),
  qty(2, "g")
)
```

# Missing data

Missing data (`NA`), empty vector (`numeric(0)`) and  infinity placeholders (`Inf`, `-Inf`) are supported in all quantities and work the same as in any other R vectors.

```{r}
qty(NA, "mg")
qty(Inf, "mg")
qty(numeric(0), "mg")
qty(c(10, NA, -Inf, Inf, numeric(0)), "mg")
```

# Miscellaneous

To check whether something is a quantity, use the `is_qty()` function. To check whether something is a specific type of quantity, use e.g. `is_mass()`.

```{r}
qty(1, "mg") %>% is_qty() # is a quantity
qty(1, "mg") %>% is_mass() # is a mass
1 %>% is_qty() # not a quantity
```

