---
title: "Quantities"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Quantities}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

All quantity data types are accessible via the universal `qty` function. The `qty` function always takes a numeric value (either single value or vector) and a unit that must match one of the supported units. All SI units and metric prefixes are supported as well as a few frequently used non-SI units (e.g. **atm** for pressure, **C** for temperature). However, non-standard units are automatically converted to the corresponding SI unit. Additionally, the `qty` function takes the optional argument `scale_to_best_metric` which determines whether the values should be automatically scaled to the most appropriate metric prefix (e.g. **10 mg** instead of **0.01 g**). By default this is done (i.e. `scale_to_best_metric = TRUE`). If the quantity is a vector, the best scaling is determined based on the median of all values.

Once familiar with the quantities supported by this package, take a look at the [operations vignette](operations.html) to see what you can do with them.

```{r}
library(mediatools) # load the library
```

## Temperature

Temperature objects are always stored in Kelvin (the SI unit) but can be generated from a number of units, the following three all generate the same temperature object.

```{r}
qty(0, "C")
qty(32, "F")
qty(273.15, "K")
```

## Amount

```{r}

```


## Mass

## Volume

## Pressure

## Molecular Weight

## Concentration



